Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-11-28 leadership_across_cultures_contexts.Rmd
Added EHB csl file
Working on introduction and results text.

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies with institutionalized responsibilities and functions associated with their position. The structure of these organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At the higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization. They are not often involved in resolving immediate conflicts, such as disputes between individuals or nested bodies, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leadership effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied of industries across 24 countries. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, leaders moderate their behavior through culturally learned values and leadership-specific values.

@@

## The benefits of leadership across cultures and contexts

@@

Institutionalized leadership positions are also often, if not universally, associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional.

Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need.

@@

@von_rueden_leadership_2015 outline four possible mechanisms by which leaders in small-scale societies might offset the potential costs associated with various leadership roles; leaders may, (1) facilitate collective actions that yield net benefits to themselves or their close kin, (2) claim a fee for their services, such a tax or greater share of returns, (3) receive reciprocal exchange in other currencies from followers who have benefited from their services, or (4) may gain other social or reproductive benefits by signaling their high quality as a mate and social partner.

## The costs of leadership across cultures and contexts

@@

@@

The most common *functions* leaders serve, which are documented in over 70% of cultures, are providing counsel or direction in follower's affairs, conflict resolution, and organizing various cooperative activities. Leaders also commonly provide social functions, serve as a representative in inter-group interactions, and provide context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punish other group members, control economic systems, and are responsible for policy development. Notable infrequent functions identified here include controlling group movement and migrations.

The most common *qualities* of leaders are being of high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent. Also, in about 70% of cultures leaders maintained coercive authority over followers. Strategic nepotism, generosity, and having a favorable personality are also common leader qualities, documented in over 50% of cultures. Notable infrequent qualities of leaders in these data include charisma and fairness.

@@

Within each cluster analyses, we computed the distance between two variables as $1 - cor (v_i, v_j)$, so that the distance between highly correlated variables is close to $0$, the distance between uncorrelated variables is close to $1$, and the distance between uncorrelated variables us close to $2$. Using the pvclust package [@suzuki_pvclust2006] we clustered the variables using the matrix of all pairwise distances using the Ward aggolomeration algorithm.

@@

The leader quality cluster analysis (Figure \@ref(fig:cluster)A) idenfitified two top-level clusters. Within each top-level cluster we identify two interpretable and generally strongly supported sub-clusters which are annotated with text. Within the first top-level cluster, we identify a dominant/shaman cluster, which suggests a strong link between dominance-based leadership (e.g., *Authority*, *Aggressiveness*, *Feared*) and shamanism (e.g., *Supernatural*, *Artistic performance*). The second sub-cluster includes qualities of leaders related to material, social, and reproductive success.

@@

A logistic PCA on the `r length(quality_vars)` coded leader quality variables, with rank $k = `r dim(logpca_model_qualities$U)[2]`$ and  $m = `r logpca_model_qualities$m`$ optimized by cross-validation, explained `r signif(logpca_model_qualities$prop_deviance_expl, 3)*100`% of the deviance. We draw on the first two leader quality components for interpretation and additional analyses. We identify the first component (PC1) as a *Prestige vs. Dominance* measure. Low values on Leader qualities PC 1 are strongly associated with the variables *Experienced/accomplished* and *Knowledgeable/intelligence* and moderately associated with *Confidence* and *Bravery*. High values are strongly associated with *Polygnous* and *Authority* and moderately associated with *High quality spouse*. Therefore, the "dominance" side of the *Prestige vs. Dominance* component is also closely liked to reproduction, as demonstrated in the cluster analysis (Figure \@ref(fig:cluster)). Measures of physical formidability and aggressiveness are also moderately associated with the "dominance" side of this component. See Figure \@ref(fig:qualitiesPCplot). The second leader qualities component we identify as an *In-group favoritism vs. Social Status* measure. Low values on the Leader qualities PC 2 are strongly associated with *Aggressiveness*, *Bravery*, *Xenophobia*, and *Loyalty*. High Values are strongly associated with *High status* and moderately associated with *Fairness* and *Age* and *Decisiveness/decision-making*.

```{r qualitiesPCplot, fig.width=15, fig.height=10, fig.cap="Logistic PCA loadings of coded leader qualitiy variables on the first two components."}
logisticPCA_loadings_plot(logpca_model_qualities, data = pca_data_qualities2)
```

Similarly, A logistic PCA on the `r length(function_vars)` coded leader function variables, with rank $k=`r dim(logpca_model_functions$U)[2]`$ and $m = `r logpca_model_functions$m`$ optimized by cross-validation, `r signif(logpca_model_functions$prop_deviance_expl, 3)*100`% of the deviance. The first Leader functions component distinguishes between social functions and general prosocial function from other more specific leader functions, which we identify as a *Social functions* component. High values on Leader function PC 1 are strongly associated with *Social functions*. The Leader functions PC 2 distinguished between organizational and mediation-based functions, which we identify as a *Organization vs. Mediation* component. Low values on Leader functions PC 2 are strongly associated with the variables *Organize cooperation* and *Political appointments*, whereas high values are strongly associated with *Resolve conflict* and moderately associated with *Provide counsel/direction*. See Figure \@ref(fig:functionsPCplot).

```{r functionsPCplot, fig.width=15, fig.height=10, fig.cap="Logistic PCA loadings of coded leader function variables on the first two components."}
logisticPCA_loadings_plot(logpca_model_functions, data = pca_data_functions2)
```

@@

Leaders are most commonly receiving benefits in the form of greater access to mateiral resources and increased social status. Receiving social services, mating benefits, and food benefits are also relatively frequent benefits of leadership cross-culturally. The costs follow the potential benefits and leaders stand to lose their high social status and greater access to resources.

Followers on the other hand also benefit from leadership systems through receiving material resources but otherwise most frequently benefit through social *services* provided. Mating and fitness benefits are relatively rare for followers. Followers most frequently incur costs in the form of material resources lost and in providing social services. For both leaders and followers, individuals stand to lose what they are able to gain. The clear differences lie in leaders' opportunities for increased social status and matting opportunities. See Figure \@ref(fig:varsupportcb).

@@

To explore potential population variation in the *Prestige vs. Dominance* leader qualities component and the *Mediation vs. Organization* leader functions component, we fit a series of hierarchical linear models at the text-level, using nested random effects of document within culture with these components as outcome measures.

>>>>

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies. Other public service and nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders CITE BASS. These positions typically are associated with institutionalized responsibilities and specific functions associated with their role. The structure of such organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At the higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization. They are not often involved in resolving immediate conflicts, such as disputes between individuals or nested bodies, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

Scholars within managerial studies have long discussed what leaders and managers do and have developed multiple models of the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [@bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group function and overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leadership functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, creating structures for accomplishing group goals, defending institutional integrity, and maintain internal consistency.

@wofford_factor_1970 provides an early empirical view of managerial functions. A factor analysis of managerial behaviors from over 130 US employees' responses to 183 questionnaire items identified five interpretable, orthogonal factors: (1) an *order and group achievement* factor, identifying participative, democratic, and organizational planning function; (2) a *personal enhancement oriented* factor, associated with authority and control, (3) a *personal interaction* factor, associated with interpersonal relationships and employee support, (4) a *security and maintenance* factor, associated with self-preservation, and (5) a *dynamic and achievement oriented* factor, associated with aggressive, self-assured actions. Wofford's analysis provided a foundation for future empirical studies of leadership and management functions. He contrasted factor one type management, described as the professional administrator who effectively collaborates and integrates team members, from factors two and five, described as autocratic leaders strongly motivated by personal power and willing to rely on aggressive tactics [@wofford_factor_1970].

More recently, many managerial scholars have adopted a comparative approach to better understand cultural diversity in leadership systems, drawing on Culturally Endorsed Implicit Leadership Theory (CLT), which suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles and behaviors [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leadership effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied of industries across 24 countries. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, leaders moderate their behavior through culturally learned values and leadership-specific values.

@@

This theoretical development ultimately produced perspectives emphasizing that organizations become larger, increasing hierarchical, adm more complex leaders... (3078 Bass) -- INTELLIGENCE and COGNITIVE COMPLEXITY IN LARGER INSTITUIONS

## The benefits and costs of leadership across cultures and contexts

@@

Institutionalized leadership positions are also often, if not universally, associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional, but is widely associated with numerous social, material, and reproductive benefits.

Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need. Across many industries and organizations leadership positions are associated with increased pay. High social status, material wealth, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006].

<!-- NEED BETTER CITATIONS ABOVE -->

@@

@von_rueden_leadership_2015 review a large body of theory and ethnography and outline four possible mechanisms by which leaders in small-scale societies might offset the potential costs associated with various leadership roles; leaders may, (1) facilitate collective actions that yield net benefits to themselves or their close kin; (2) claim a fee for their services, such a tax or greater share of returns; (3) receive reciprocal exchange in other currencies from followers who have benefited from their services; or (4) may gain other social or reproductive benefits by signaling their high quality as a mate and social partner.

@@

Conceptualizing leadership as a possible solution to problems associated with collective goods [@olson_logic_1965], where group members pool resources and share costs of public goods to receive a net benefit from a cooperative endeavor, leader incur costs associated with managing contributions, coordinating behaviors, and broadcasting information to maintain coordination and cooperation. In this framework, leaders and followers voluntary participate in a cooperative enterprise with the expectation that benefits will outweigh the imposed costs. @rich_dynamics_1980 adopts this framework of collective costs and benefits to understand leadership and group dynamics in urban neighborhoods and neighborhood organizations, which can be viewed as self-interested groups which face group-wide problems and necessarily share public benefits. The costs associated with leadership in neighborhood organizations follow those associated with other collective enterprises (e.g., investments of time and energy in management and maintenance of group coordination). @rich_dynamics_1980 conceptualizes four categories of the benefits associated with neighborhood organization. Leaders of neighborhood organizations (1) may receive benefits of the public goods produces from collective actions (as other non-leaders would as well); (2), may receive selective benefits from surpluses produced through collective actions; (3) may receive incidental benefits from their leadership role; and (4) leaders of neighborhood organizations may enjoy their service in the position. @rich_dynamics_1980 interviewed 54 officers from 11 metropolitan neighborhood organizations in the Midwest US. Results from extensive interviews suggest personal fulfillment and enjoyment are the main motivational drivers of participation in leadership roles of neighborhood organizations, suggesting the process of leadership for individuals is distinct from theoretical models rooted in economics. Therefore, a primary benefit of leadership, in some contexts, is likely to be intrinsic personal fulfillment and the opportunity to execute personal values of community service.

@@

The most common *functions* leaders serve, which are documented in over 70% of cultures, are resolving conflicts, providing various social functions, providing counsel or direction in follower's affairs, and organizing various cooperative activities. Leaders also commonly serve as a representative in inter-group interactions, manage the distribution of resources, and provide context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punish other group members and manage economic systems. Notable infrequent functions identified here include controlling group movement and migrations.

The most common *qualities* of leaders are being of high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of the cultures sampled here leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, and maintained strong authority over followers. Notable infrequent qualities of leaders in these data include charisma and fairness.

@@

Within each cluster analyses, we computed the distance between two variables as $1 - cor (v_i, v_j)$, so that the distance between highly correlated variables is close to $0$, the distance between uncorrelated variables is close to $1$, and the distance between uncorrelated variables us close to $2$. Using the pvclust package [@suzuki_pvclust2006] we clustered the variables using the matrix of all pairwise distances using the Ward agglomeration algorithm.

@@

The leader quality cluster analysis (Figure \@ref(fig:cluster)A) identified two top-level clusters. Within each top-level cluster we identify two interpretable and generally strongly supported sub-clusters which are annotated with text. Within the first top-level cluster, we identify a dominant/shaman cluster, which suggests a strong link between dominance-based leadership (i.e., *Authority*, *Aggressiveness*, *Feared*) and shamanism (i.e., *Supernatural*, *Artistic performance*). The second sub-cluster includes qualities of leaders related to material, social, and reproductive success (i.e., *Wealthy*, *Social contacts*, and *Polygnous*). Within the second top-level, we identify a sub-cluster of positive personal characteristics (i.e., *Knowledgeable/intelligent*, *Proper behavior*, *Charisma*, *Fairness*) and a sub-cluster of measures related to social strategies and health (i.e., *Strategic planner*, *Generosity*, *Physically formidable*).

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identify as an organizational functions cluster (i.e., *Organize cooperation*, *Military command*, *Resolve conflict*) and the smaller a prosociality cluster (i.e., *Social functions*, *Distribute resources*, *Hospitality*).

@@

A logistic PCA on the `r length(quality_vars)` coded leader quality variables, with rank $k = `r dim(logpca_model_qualities$U)[2]`$ and  $m = `r logpca_model_qualities$m`$ optimized by cross-validation, explained `r signif(logpca_model_qualities$prop_deviance_expl, 3)*100`% of the deviance. We draw on the first two leader quality components for interpretation and additional analyses (see Figure \@ref(fig:qualitiesPCplot)). We identify the first component (PC1) as a *Prestige vs. Dominance* measure. Low values on Leader Qualities PC 1 are strongly associated with the variables *Experienced/accomplished* and *Knowledgeable/intelligence* and moderately associated with *Confidence* and *Bravery*. High values are strongly associated with *Polygnous* and *Authority* and moderately associated with *High quality spouse*. Therefore, the "dominance" side of the *Prestige vs. Dominance* component is also closely liked to reproduction, as demonstrated in the cluster analysis (Figure \@ref(fig:cluster)). Measures of physical formidability and aggressiveness are also moderately associated with the "dominance" side of this component. The second leader qualities component we identify as an *In-group favoritism vs. Social Status* measure. Low values on Leader Qualities PC 2 are strongly associated with *Aggressiveness*, *Bravery*, *Xenophobia*, and *Loyalty*. High Values are strongly associated with *High status* and moderately associated with *Fairness*, *Age*, and *Decisiveness/decision-making*. Therefore, Leader Qualities PC 2 also reflects elements of a presetige-dominance distinction in contrasting aggressiveness with high social status.

```{r qualitiesPCplot, fig.width=15, fig.height=10, fig.cap="Logistic PCA loadings of coded leader qualitiy variables on the first two components. PC 1 is identified as a *Prestige vs. Dominance* component and PC 2 is identified as an *In-group favoritism vs. Social Status* component."}
logisticPCA_loadings_plot(logpca_model_qualities, data = pca_data_qualities2)
```

Similarly, A logistic PCA on the `r length(function_vars)` coded leader function variables, with rank $k=`r dim(logpca_model_functions$U)[2]`$ and $m = `r logpca_model_functions$m`$ optimized by cross-validation, explained `r signif(logpca_model_functions$prop_deviance_expl, 3)*100`% of the deviance. The first leader functions component distinguishes between social functions (a very broad and generalized measure) from other more specific leader functions, which we identify as a *Social functions* component. Low values on Leader function PC 1 are strongly associated with *Social functions*. The Leader functions PC 2 distinguished between organizational and mediation-based functions, which we identify as an *Organization vs. Mediation* component. Low values on Leader functions PC 2 are strongly associated with the variables *Organize cooperation* and *Political appointments*, whereas high values are strongly associated with *Resolve conflict* and moderately associated with *Provide counsel/direction*. See Figure \@ref(fig:functionsPCplot).

```{r functionsPCplot, fig.width=15, fig.height=10, fig.cap="Logistic PCA loadings of coded leader function variables on the first two components. PC 1 is identified as a *Social functions* component and PC 2 is identified as an *Organization vs. Mediation* component."}
logisticPCA_loadings_plot(logpca_model_functions, data = pca_data_functions2)
```

@@

In these data leaders are most commonly receiving benefits in the form of greater access to material resources (documented in about 70% of cultures) and increased social status (documented in about 65% of cultures). Receiving social services, mating benefits, and food benefits are also relatively frequent benefits of leadership cross-culturally. The costs follow the potential benefits and leaders stand to lose their high social status (documented in about 45% of cultures) and greater access to resources (documented in about 40% of cultures).

Followers on the other hand also benefit from leadership systems through receiving material resources (documented in about 50% of cultures) but otherwise most frequently benefit through social *services* provided (documented in about 35% of cultures). Mating and fitness benefits are relatively rare for followers. Followers most frequently incur costs in the form of material resources lost (documented in about 40% of cultures) and in providing social services (documented in about 25% of cultures).

For both leaders and followers, individuals stand to lose what they are able to gain. The clear differences lie in leaders' opportunities for increased social status and matting opportunities. See Figure \@ref(fig:varsupportcb).

@@

To explore potential population variation in the *Prestige vs. Dominance* and *In-group favoritism vs. Social Status* leader qualities components and the *Social functions* and *Mediation vs. Organization* leader functions components, we fit a series of hierarchical linear models at the text-level, using nested random effects of document within culture with these components as outcome measures.
